Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-20 background.tex
+

In Figure \ref{fig:pagetablescode}, we see a simplified kernel code, \lstinline|walkpgdir|, implementing this page-table walk for translating the virtual address $\vaddr$ over the page-table tree with unique root address $\mathsf{root}$ shown in Figure \ref{fig:pagetables}.
As shown in Figure at the top of Figure \ref{fig:pagetables}, on x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 
1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits 
being used to index into this page.\footnote{Technically levels 1--3 have explicit historical names,
but for brevity and consistency, we simply number them, in keeping with the newer 5th level. Our formalization only deals with 4-level page tables, but is straightforwardly extensible to 5.} 
As depicted at the top of Figure \ref{fig:pagetables}, for each level of translation, address-translation implementation repeatedly takes designated slices of the virtual address and indexing into a table --- e.g., L4Offset(va) at Line 2 of \lstlinline|walkpgdir| in Figure \ref{fig:pagetablescode}. 
The final lookup in the page tables gives the base physical address of a 4KB page of physical memory, to which the low-order
bits of the accessed virtual address are added to determine the actual physical address retrieved. The entries of each table are 64 bits wide, but each points to a physical address aligned to 4KB (4096 byte) boundaries, which leaves 12 bits to spare to 
control a validity bit (called the \emph{present} bit), a read-write bit (which permits write access through the entry if and only if it is set), and a range of additional bits which can be used to control caching, write-through, and more. This paper will only consider the present bit (0).

>>>>

In Figure \ref{fig:pagetablescode}, we see a simplified kernel code, \lstinline|walkpgdir|, implementing this page-table walk for translating the virtual address $\vaddr$ over the page-table tree with unique root address $\mathsf{root}$ shown in Figure \ref{fig:pagetables}.
As shown in Figure at the top of Figure \ref{fig:pagetables}, on x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 
1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits 
being used to index into this page.\footnote{Technically levels 1--3 have explicit historical names,
but for brevity and consistency, we simply number them, in keeping with the newer 5th level. Our formalization only deals with 4-level page tables, but is straightforwardly extensible to 5.} 
As depicted at the top of Figure \ref{fig:pagetables}, for each level of translation, address-translation implementation repeatedly takes designated slices of the virtual address and indexing into a table --- e.g., L4Offset(va) at Line 2 of \lstlinline|walkpgdir| in Figure \ref{fig:pagetablescode}. 
The final lookup in the page tables gives the base physical address of a 4KB page of physical memory, to which the low-order
bits of the accessed virtual address are added to determine the actual physical address retrieved. The entries (e.g., L4 Entry in Figure \ref{fig:pagetables}) of each table are 64 bits wide, but each points to a physical address aligned to 4KB (4096 byte) boundaries, which leaves 12 bits to spare to 
control a validity bit (called the \emph{present} bit), a read-write bit (which permits write access through the entry if and only if it is set), and a range of additional bits which can be used to control caching, write-through, and more. This paper will only consider the present bit (0). As we see in Figure \ref{fig:pagetables}, the entry L2 is not be present (shown as dashed in Figure \ref{fig:pagetables}), ...
